The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
restructuring

@@

A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge}.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
The vertex set of the \emph{$\rnni$-graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space}.

@@

\section{Caterpillar Trees}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
	\caption{Paths between caterpillar trees $T_1$ and $T_2$: black -- $\nni$ path of length $5$; red -- $\rnni$ path of length $6$; blue -- $\rnni$ path of length $8$.}
	\label{NNI_vs_RNNI}
\end{figure}

\begin{conjecture}
	The set of caterpillar trees is convex in $\rnni$ space.
\end{conjecture}

This conjecture is computationally proven for up to $8$ taxa (maximum $\rnni$ graph we can compute).
For a more restricted subset of trees in $\rnni$, namely caterpillar trees that have one cherry leaf in common, one can show convexity.

\begin{lemma}
	Let $T_1$ and $T_2$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
	There is a shortest paths from $T_1$ to $T_2$ that is a caterpillar path.
	\label{lemma:caterpillar_subset_convex}
\end{lemma}

\begin{figure}[h]
	\centering
	\includegraphics[width=0.3\textwidth]{label_internal_nodes}
	\caption{The labelling of the caterpillar tree as introduced in the proof of Lemma~\ref{lemma:caterpillar_subset_convex} is unique due to its definition.
	}
	\label{label_internal_nodes}
\end{figure}

\begin{proof}
	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
	Let $p$ be a shortest path from $T_1$ to $T_2$.
	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.

	In $T_1$ and $T_2$ the internal node of the cherry is labelled by the label of its child that is not $a$.
	All other internal nodes shall have the same label as their leaf children.
	From now on we will identify the leaf labelled by $x$ with $x$ and the internal node labelled by $x$ will be called $int_x$.
	An example for a caterpillar tree with internal leaf labels is depicted in Figure~\ref{label_internal_nodes}.
	In the following, changes of the internal labelling following $\rnni$ moves are defined:

	\textbf{Rank moves:}
	Within rank moves the labels of internal nodes do not change.
	For an example consider Figure~\ref{fig:rank_swap_internal_label}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank move.
		}
		\label{fig:rank_swap_internal_label}
	\end{figure}

	\textbf{NNI move:}
	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $rank(int_x) < rank(int_y)$ as in Figure~\ref{fig:nni_moves}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.7\textwidth]{nni_moves_internal_label}
		\caption{Two possible $\nni$ moves on edge $(int_x, int_y)$ where $x \in A, y \in B$.
		}
		\label{fig:nni_moves}
	\end{figure}

	The nodes incident to the edge after the $\nni$ move are labelled by $int_x$ and $int_y$ as well.
	By default, the node that has lower rank than the other one is labelled by $int_x$, the other one by $int_y$.
	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $rank(int_x) > rank(int_y)$.
	An example of an $\nni$ move and the new labelling is provided in Figure~\ref{fig:nni_example_internal_label}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\textwidth]{nni_example_internal_label}
		\caption{The first labelling after the $\nni$ move preserves the ranks of $int_4$ and $int_6$.
		But then $int_6$ is not an ancestor of $6$ any more.
		Therefore, $int_4$ and $int_6$ have to change their labels.
		}
		\label{fig:nni_example_internal_label}
	\end{figure}

	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
	If $x$ stays in the subtree rooted in the lower node of $e$, this does obviously hold.
	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.
	This makes sure that $int_x$ is an ancestor of $x$.
	Thus, it remains to prove that $int_y$ is ancestor of $y$.
	In particular, we need to show that $y$ is not in the same subtree $A$ as $x$, whose root has the node labelled by $int_x$ as parent before and after the $\nni$ move.
	Let us consider the tree before the $\nni$ move.
	$A$ has $|A|$ leaves and $|A| - 1$ internal nodes.
	These internal nodes are labelled by some leaves of $A$, because all internal nodes must be labelled by a descendant leaf.
	We know that $x \in A$, and the parent of the root of $A$ is labelled by $x$.
	All remaining $|A|-1$ leaf labels of $A$ appear as labels of the $|A|-1$ internal nodes of $A$.
	From the fact that the internal node labelled by $y$ is not inside $A$ we can conclude that $y$ cannot be a leaf of $A$.
	This proves that the labelling of the internal nodes is is well defined and allows a unique identification of internal nodes by their labels.

	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
	Therefore, we use the following observation:
	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
	This is true for both rank moves and $\nni$ moves, according to our definition of the labelling:
	Within a rank move two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.

	On a caterpillar path the only possible moves are $\nni$ moves that change the ranks of two internal nodes by one each.
	Therefore, we can construct a caterpillar path $r$ out of $p$:
	every time two internal nodes $int_x$ and $int_y$ exchange ranks on $p$, they do so on the caterpillar path $r$ as well.
	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
\end{proof}

\section{Split Theorem}

@@

\section{Algorithms}

We are now going to introduce different algorithms that computes paths in $\rnni$.
\cite{Gavryushkin2017} gave an algorithm for computing the whole $\rnni$ space.
We implemented this algorithm and are able to use it on trees with up to $8$ taxa.
Therefore, we can compare the paths given by the algorithms presented in this chapter with the exact shortest path for trees with up to $8$ taxa.

% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.

A modified version of the classic Bubble Sort \todo{cite} algorithm returns a caterpillar path that is shortest among all caterpillar paths between given caterpillar trees $T_1, T_2$.
Let $T_1 = (\ldots((1,2)3) \ldots )n), T_2 = (\ldots((\sigma_1,\sigma_2)\sigma_3) \ldots )\sigma_n)$ be two caterpillar trees.
They can be seen as sequences $(1,2,\ldots,n)$ and $(\sigma_1, \sigma_2, \ldots, \sigma_n)$ that need to be sorted.
$\nni$ moves between caterpillar trees correspond to exchanging two neighboured elements in a sequence.
The only exception of this are $\nni$ moves that change the cherry of a tree.
The corresponding moves on sequences are exchanged of either the second and third element or the first and third element.
On the other hand, there is no difference between two the trees corresponding to two sequences where only the first two taxa are exchanged.
Since all pairs of taxa whose order in $T_1$ is different from the one in $T_2$ need to be exchanged by $\nni$ moves on a caterpillar path, Bubble Sort will give a shortest path among all caterpillar paths.
Just the moves for the first three positions of the sequence must be adapted.
Let us consider the first three elements $(a,b,c,\ldots)$ of a sequence representing a caterpillar tree on a caterpillar path between $T_1$ and $T_2$
If the order of these elements is the same in $T_2$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $T_2$, this one will change wih $c$.
But if both $a$ and $b$ are on the other side of $c$ in $T_2$, the one which is the righmost of all three taxa in $T_2$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $T_2$.
\todo{This paragraph is just a draft. Is it needed at all?}

Following algorithm computes paths between two trees in $\rnni$:

\begin{algorithm}[H]
\caption{FIND\_PATH($T_1,T_2$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $T := T_1$
	\STATE $p := (T_1)$
	\FOR {$r = 1, \dots, n-2$}
		\STATE Let $S$ be the cluster associated with the node $v$ in $T_2$ with $rank(v) = r$
		\WHILE {$rank(MRCA_T(S))>r$}
			\STATE Update $T$: Decrease $rank(MRCA_T(S))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+T$
		\ENDWHILE
	\ENDFOR
	\RETURN p
\end{algorithmic}
\end{algorithm}

One can easily prove that there is always exactly one possible $\rnni$ move that decreases the rank of $MRCA_T(S)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.

The proof of Theorem 7 in \cite{Gavryushkin2017} gives an algorithm to compute paths between trees in $\rnni$. %Do we want to mention that algorithm here?

% Compare the two algorithms by sampling trees and check which algorithm gives shorter paths.

% Also mention the exact algorithm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
% Maybe it's worth trying the other algorithm in these trees as well.

Algorithm~\ref{alg:find_path} provides an upper bound for the diameter of $\rnniu$ space.
It improves the bound $n^2 - 3n - \frac{5}{8}$ given in \cite{Gavryushkin2017}.

\begin{theorem}
	It is $\Delta(\rnniu) \leq \frac{1}{2}n^2-\frac{3}{2}n+1$
\end{theorem}

\begin{proof}
	This result directly follows from Algorithm~\ref{alg:find_path}.
	In the worst case each $r = 1, \dots, n-2$ requires $n-1-r$ $\rnni$ moves for moving the corresponding $MRCA$ down.
	Hence, there is for each pair of trees a path of length less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2} = \frac{1}{2}n^2-\frac{3}{2}n+1$.
\end{proof}

\begin{lemma}
	There are two caterpillar trees in $\rnniu$ with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$.
	\label{conj:caterpillar_diameter}
\end{lemma}

\begin{proof}
    For this proof we will give two caterpillar trees $T_1$ and $T_2$ on $n$ taxa with $d_{RNNI}(T_1,T_2) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
    \todo{$d$ or $d_{RNNI}$?}
	Let $T_1 = (( \dots (n,1)2)\dots)n-1)$ and $T_2 = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Lemma~\ref{lemma:caterpillar_subset_convex}, there is a shortest path between $T_1$ and $T_2$ that only consists of caterpillar trees.
    We can compute such a shortest caterpillar path by using a modified version of Bubble Sort.
    When running this algorithm on $T_1$ and $T_2$ there are $n-1-i$ swaps of taxon $i$ with its right neighbour.
    Therefore, the total number of these swaps, and therefore $d_{RNNI}(T_1,T_2)$, is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{1}{2}n^2-\frac{3}{2}n+1$
\end{proof}

\begin{corollary}
It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2} = \frac{1}{2}n^2-\frac{3}{2}n+1$.
\end{corollary}

%This lemma might be central  for induction proofs in RNNI space
%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
\begin{lemma}
    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{theorem}
    The set of caterpillar trees is convex.
\end{theorem}

% With the knowledge that the set of caterpillar trees is convex in $\rnniu$, we can prove the Cluster Theorem restricted to caterpillar trees.

\begin{conjecture}
    Let $T_1$ and $T_2$ be caterpillar trees.
    If $T_1$ and $T_2$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T_1$ and $T_2$. \todo{Cluster means non-trivial cluster -> def chapter?}
\end{conjecture}

% Following are some ideas that did not lead to major results so far:
% idea for description of max distance caterpillar trees:

@@

\section{Partition lattice}

>>>>

%Introduce RNNI as ranked nearest neighbour interchange space, only ultrametric trees considered (extension could be mentioned in the end)

%We use the exact algorithm to calculate all the distances for the examples given in this paper

@@

A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if ${|rank(v) - rank(u)| = l}$.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
The vertex set of the \emph{$\rnni$-graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space} or just $\rnni$.

@@

\section{Geometry of $\rnni$}

The following Algorithm~\ref{alg:find_path} which computes paths between pairs of trees in $\rnni$ is a central tool for our investigation of shortest paths in $\rnni$.

\begin{algorithm}[H]
\caption{FIND\_PATH($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with ${rank(v) = i < rank(w)}$
		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

\todo{figure of an example?}

One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in~\cite{Gavryushkin2017}:

\begin{lemma}
    It is $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$.
\end{lemma}

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ given by internal edges of $R$, represented by $i = 1, \dots, n-2$ in the algorithm, requires $n-1-i$ $\rnni$ moves for moving $mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to ${\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}}$.
\end{proof}

Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following induction proofs. \todo{Is it correct and if it is, do we need to prove the statement, or is it rather obvious?}

\todo{Change notations in this lemma+proof (less indices!!)}
\begin{lemma}
    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\subsection{The set of caterpillar trees}
\label{section:caterpillar_convex}

The set of caterpillar trees is of interested in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
Since $\nni$ moves in $\rnni$ are only allowed on edges of length one, the structure of shortest paths between caterpillar trees is different in $\nni$ and $\rnni$ as it is pointed out in Figure~\ref{fig:NNI_vs_RNNI}.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
	\caption{Paths between caterpillar trees $T$ and $R$: black -- shortest path in classical $\nni$ (ignoring ranks) of length $5$; red -- shortest path in $\rnni$ of length $6$; blue -- path in $\rnni$ which is the natural extension of the shortest path in $\nni$ but has length $8$.}
	\label{fig:NNI_vs_RNNI}
\end{figure}

This observation already suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
Indeed, our aim for this section is to prove Theorem~\ref{thm:caterpillar_convex} which states that the set of caterpillar trees is convex in $\rnni$.
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.
One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between caterpillar trees, coincide with exchanges of two neighboured elements in the sequence.
In these sequences the first two taxa are somehow equivalent since they build the unique cherry.
$\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm~\cite{knuth1997}:

Following the observations above, only the moves for the first three positions of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $parent(c)$ is higher than the ranks of $parent(a)$ and $parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.

\begin{theorem}
    The set of caterpillar trees is convex.
    \label{thm:caterpillar_convex}
\end{theorem}

\subsection{Diameter}

In~\cite{Gavryushkin2017} upper and lower bound for the diameter of $\rnni$ are given.
Following the results of section~\ref{section:caterpillar_convex} one can directly improve this result by giving the exact diameter of the $\rnni$ graph.

\begin{lemma}
	There are two caterpillar trees in $\rnniu$ with distance $\frac{(n-1)(n-2))}{2}$.
	\label{conj:caterpillar_diameter}
\end{lemma}

\begin{proof}
	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
    We can compute a shortest caterpillar path by using the modified version of Bubble Sort as explained above. \todo{Do we want to name this algorithm?}
    When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
    Therefore, the length of a shortest caterpillar path is ${\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}}$.
\end{proof}

\begin{corollary}
    It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
\end{corollary}

%Do we need this?

% \begin{lemma}
% 	Let $T_1$ and $T_2$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
% 	There is a shortest paths from $T_1$ to $T_2$ that is a caterpillar path.
% 	\label{lemma:caterpillar_subset_convex}
% \end{lemma}
%
% \begin{figure}[h]
% 	\centering
% 	\includegraphics[width=0.3\textwidth]{label_internal_nodes}
% 	\caption{The labelling of the caterpillar tree as introduced in the proof of Lemma~\ref{lemma:caterpillar_subset_convex} is unique due to its definition.
% 	}
% 	\label{label_internal_nodes}
% \end{figure}
%
% \begin{proof}
% 	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
% 	Let $p$ be a shortest path from $T_1$ to $T_2$.
% 	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.
%
% 	In $T_1$ and $T_2$ the internal node of the cherry is labelled by the label of its child that is not $a$.
% 	All other internal nodes shall have the same label as their leaf children.
% 	From now on we will identify the leaf labelled by $x$ with $x$ and the internal node labelled by $x$ will be called $int_x$.
% 	An example for a caterpillar tree with internal leaf labels is depicted in Figure~\ref{label_internal_nodes}.
% 	In the following, changes of the internal labelling following $\rnni$ moves are defined:
%
% 	\textbf{Rank moves:}
% 	Within rank moves the labels of internal nodes do not change.
% 	For an example consider Figure~\ref{fig:rank_swap_internal_label}.
%
% 	\begin{figure}[H]
% 		\centering
% 		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
% 		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank move.
% 		}
% 		\label{fig:rank_swap_internal_label}
% 	\end{figure}
%
% 	\textbf{NNI move:}
% 	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $rank(int_x) < rank(int_y)$ as in Figure~\ref{fig:nni_moves}.
%
% 	\begin{figure}[H]
% 		\centering
% 		\includegraphics[width=0.7\textwidth]{nni_moves_internal_label}
% 		\caption{Two possible $\nni$ moves on edge $(int_x, int_y)$ where $x \in A, y \in B$.
% 		}
% 		\label{fig:nni_moves}
% 	\end{figure}
%
% 	The nodes incident to the edge after the $\nni$ move are labelled by $int_x$ and $int_y$ as well.
% 	By default, the node that has lower rank than the other one is labelled by $int_x$, the other one by $int_y$.
% 	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $rank(int_x) > rank(int_y)$.
% 	An example of an $\nni$ move and the new labelling is provided in Figure~\ref{fig:nni_example_internal_label}.
%
% 	\begin{figure}[H]
% 		\centering
% 		\includegraphics[width=0.8\textwidth]{nni_example_internal_label}
% 		\caption{The first labelling after the $\nni$ move preserves the ranks of $int_4$ and $int_6$.
% 		But then $int_6$ is not an ancestor of $6$ any more.
% 		Therefore, $int_4$ and $int_6$ have to change their labels.
% 		}
% 		\label{fig:nni_example_internal_label}
% 	\end{figure}
%
% 	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
% 	If $x$ stays in the subtree rooted in the lower node of $e$, this does obviously hold.
% 	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
% 	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.
% 	This makes sure that $int_x$ is an ancestor of $x$.
% 	Thus, it remains to prove that $int_y$ is ancestor of $y$.
% 	In particular, we need to show that $y$ is not in the same subtree $A$ as $x$, whose root has the node labelled by $int_x$ as parent before and after the $\nni$ move.
% 	Let us consider the tree before the $\nni$ move.
% 	$A$ has $|A|$ leaves and $|A| - 1$ internal nodes.
% 	These internal nodes are labelled by some leaves of $A$, because all internal nodes must be labelled by a descendant leaf.
% 	We know that $x \in A$, and the parent of the root of $A$ is labelled by $x$.
% 	All remaining $|A|-1$ leaf labels of $A$ appear as labels of the $|A|-1$ internal nodes of $A$.
% 	From the fact that the internal node labelled by $y$ is not inside $A$ we can conclude that $y$ cannot be a leaf of $A$.
% 	This proves that the labelling of the internal nodes is is well defined and allows a unique identification of internal nodes by their labels.
%
% 	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
% 	Therefore, we use the following observation:
% 	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
% 	This is true for both rank moves and $\nni$ moves, according to our definition of the labelling:
% 	Within a rank move two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.
%
% 	On a caterpillar path the only possible moves are $\nni$ moves that change the ranks of two internal nodes by one each.
% 	Therefore, we can construct a caterpillar path $r$ out of $p$:
% 	every time two internal nodes $int_x$ and $int_y$ exchange ranks on $p$, they do so on the caterpillar path $r$ as well.
% 	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
% 	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
% \end{proof}

\subsection{Split Theorem}

@@

\section{Ideas}
\todo{Which ideas do we want to keep? Shall this become an 'open problems/future work' section?}

Following are some ideas that did not lead to major results so far:
% idea for description of max distance caterpillar trees:

@@

\subsection{Partition lattice}
